# Recap on Linear Models

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false
library(dplyr)
library(readr)
library(ggplot2)
source("ggplot2_setup.R", local=TRUE)
set.seed(332548978)  # from sample(.Machine$integer.max, 1)
```

## Building up the Linear Model

A common behavioural task in language and word recognition research is the *Lexical Decision Task*. In Lexical Decision Tasks, we present real words (e.g., table) and word-like pseudowords (e.g., frimp). We then measure people's responses and their response times to each item.

The best predictor of response time to real words is word frequency: response times are faster to words that occur more often.

Here are some made-up data showing the relationship between *Zipf word frequency* [a log-transformed count variable; @vanheuvenSUBTLEXUKNewImproved2014] and *response time* (ms). Each point is depicts one word.

```{r}
#| label: sim_dat
#| echo: false
#| message: false
#| warning: false
#| fig-height: 4
#| fig-width: 4
b0 <- 6.95
b1 <- -0.11
s <- 0.12
n <- 100

dat <- tibble(
  zipf = runif(n, 1.3, 7.6),
  e_i = rnorm(n, 0, s),
  log_rt = b0 + b1 * zipf + e_i
)

write_csv(dat, file.path("docs", "sims_lm.csv"))

dat |>
  ggplot(aes(zipf, log_rt)) +
  geom_point() +
  labs(x="Word Frequency (Zipf)", y="Log Response Time (ms)")
```

How could we describe these data?

### 1. The Average (Intercept)

One very simple but powerful description could be to just calculate the average log response time.

```{r}
mean(dat$log_rt)
```

# Intercept and Slope

$$\begin{aligned}
& y_i = \beta_0 + \beta_1 x_i + e_i \\
& e_i \sim N(0, \sigma)
\end{aligned}$$

```{r}
1 + 1
```
