# LMMs: Random Intercepts

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false
source("ggplot2_setup.R", local=TRUE)
set.seed(1019162963)  # from sample(.Machine$integer.max, 1)
```

## Why Linear Mixed-Effects Models?

In the section recapping [linear models](01-linear_models), we used a linear model to describe the relationship between word frequency and log response time in a lexical decision task. In these data we had one observation per word.

In a real study, however, we would have multiple observations for each word from each participant. The traditional approach would be to *aggregate*: calculate the average log response time for each word before fitting the model. However, while this allows us to account for between-item variability, by aggregating over participants, we are losing valuable information about how the data were generated.

For some time, a common compromise was to fit the model twice, and then examine whether a hypothesis is supported across both items and participants [@raaijmakersHowDealLanguageasFixedEffect1999]. However, a more robust solution would be to model item and participant variability simultaneously. This is something that linear-mixed effects models allow us to do [@lockerUseMultilevelModeling2007].

## What is a Linear Mixed-Effects Model?

A linear mixed-effects model is largely a generalisation of the ideas behind a linear model that we went over in the [recap on linear models](01-linear_models). A key difference is that rather than aggregating, we can model multiple sources of variability simultaneously.

To start with, we will just work with random intercepts:

$$logRT_{ip} = \beta_0 + \beta_1 Zipf_{ip} + I0_i + P0_p + e_{ip}$$
Our formula is very similar to before, with the following key differences:

* We index observations with two indices: $i$ (item) and $p$ (participant).
* We include $I0_i$: per-item random intercepts
* We include $P0_p$: per-participant random intercepts

As we saw with [$e_i$ in the linear model example](01-linear_models.html#intercept-slope-residual), $I0_i$, $P0_p$, and $e_{ip}$ are all drawn from normal distributions with their own standard deviations:

$$\begin{aligned}
I0_i &\sim N(0, \sigma_{I0}) \\
P0_p &\sim N(0, \sigma_{P0}) \\
e_{ip} &\sim N(0, \sigma_e)
\end{aligned}$$

To get a sense for what this looks like, in this section we'll use these formulae to simulate some data.

## Packages

We'll use the following packages in this section:

```{r}
#| label: packages
#| message: false
#| warning: false
library(lme4)  # for fitting LMMs
library(dplyr)  # for data wrangling
library(readr)  # for importing data
library(ggplot2)  # for plotting
```

## Parameters

Here we define all the free parameters from the equations above:

* $\beta_0$: `b_0`
* $\beta_1$: `b_1`
* $\sigma_{I0}$: `s_P0`
* $\sigma_{P0}$: `s_I0`
* $\sigma_e$: `s_e`

The numbers here are based on the same real data as in the [section on linear models](01-linear_models).

```{r}
#| label: params
# Fixed-effect intercept and slope
# (same as the parameters in the section on linear models)
b_0 <- 6.85
b_1 <- -0.1

# SD of per-item random intercepts
s_I0 <- 0.08

# SD of per-participant random intercepts
s_P0 <- 0.16

# residual SD
s_e <- 0.33
```


