# LMMs: Random Intercepts

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false
source("ggplot2_setup.R", local=TRUE)
set.seed(1019162963)  # from sample(.Machine$integer.max, 1)
```

## Why Linear Mixed-Effects Models?

In the section recapping [linear models](01-linear_models), we used a linear model to describe the relationship between word frequency and log response time in a lexical decision task. In these data we had one observation per word.

In a real study, however, we would have multiple observations for each word from each participant. The traditional approach would be to *aggregate*: calculate the average log response time for each word before fitting the model. However, while this allows us to account for between-item variability, by aggregating over participants, we are losing valuable information about how the data were generated.

Where it was possible, a common compromise could be to fit the model twice, and then examine whether a hypothesis is supported across both items and participants [@raaijmakersHowDealLanguageasFixedEffect1999]. However, a more robust solution would be to model item and participant variability simultaneously. This is something that linear-mixed effects models allow us to do [@lockerUseMultilevelModeling2007].

## What is a Linear Mixed-Effects Model?

A linear mixed-effects model is largely a generalisation of the ideas behind a linear model that we went over in the [recap on linear models](01-linear_models). A key difference is that rather than aggregating, we can model multiple sources of variability simultaneously.

To start with, we will just work with random intercepts:

$$logRT_{ip} = \beta_0 + \beta_1 Zipf_{ip} + I0_i + P0_p + e_{ip}$$
Our formula is very similar to before, with the following key differences:

* We index observations with two indices: $i$ (item) and $p$ (participant).
* We include $I0_i$: per-item random intercepts
* We include $P0_p$: per-participant random intercepts

As we saw with [$e_i$ in the linear model example](01-linear_models.html#intercept-slope-residual), $I0_i$, $P0_p$, and $e_{ip}$ are all drawn from normal distributions with their own standard deviations:

$$\begin{aligned}
I0_i &\sim N(0, \sigma_{I0}) \\
P0_p &\sim N(0, \sigma_{P0}) \\
e_{ip} &\sim N(0, \sigma_e)
\end{aligned}$$

The per-item and per-participant random intercepts describe variability among these populations. Items will vary in their average response time --- while word frequency is maybe the best predictor of response times, a whole range of other factors will also affect a single word's average response time. Similarly, participants will differ in their average response times, with some participants responding consistently faster, and some responding consistently slower. Our model says that these random effects combine with our fixed effects to generate our observations.

To get a sense for what this looks like, in this section we'll use these formulae to simulate some data.

## Packages

We'll use the following packages in this section:

```{r}
#| label: packages
#| message: false
#| warning: false
library(lme4)  # for fitting LMMs
library(dplyr)  # for data wrangling
library(tidyr)  # for data wrangling
library(readr)  # for importing data
library(ggplot2)  # for plotting
```

## Defining Parameters

Here we define all the free parameters from the equations above:

* $\beta_0$ (Fixed-effect Intercept): `b_0`
* $\beta_1$ (Fixed-effect Slope): `b_1`
* $\sigma_{I0}$ (SD of per-item random intercepts): `s_P0`
* $\sigma_{P0}$ (SD of per-participant random intercepts): `s_I0`
* $\sigma_e$ (SD of the residual): `s_e`

The numbers here are based on the same real data as in the [section on linear models](01-linear_models).

```{r}
#| label: params
# Fixed-effect intercept and slope
# (same as the parameters in the section on linear models)
b_0 <- 6.85
b_1 <- -0.1

# SD of per-item random intercepts
s_I0 <- 0.08

# SD of per-participant random intercepts
s_P0 <- 0.16

# residual SD
s_e <- 0.33
```

## Simulating Items and Participants

To be able to simulate trials (a single participant responds to a single item), we first need to simulate the individual items and participants.

Here we build two dataframes, of 50 items and 30 participants.

```{r}
# simulate 50 items
n_items <- 50

items <- tibble(
  # item IDs
  i_id = 1:n_items,
  # item word frequencies
  zipf = runif(n_items, 1.5, 7.5),
  # item intercepts
  I0_i = rnorm(n_items, 0, s_I0)
)

# simulate 30 participants
n_participants <- 30

participants <- tibble(
  # participant IDs
  p_id = 1:n_participants,
  # participant intercepts
  P0_p = rnorm(n_participants, 0, s_P0)
)
```

## Visualising Random Intercepts

As you can see, the random intercepts allow us to describe multiple sources of variability at the same time. A common way of understanding random effects is to plot separate linear relationships for each unit in our population.

For example, just as we can [plot a single simulated linear relationship](01-linear_models.html#intercept-slope), so too can we plot how this would look for items and participants in a given sample:

```{r}
#| layout-ncol: 2
#| fig-height: 4
#| fig-width: 3
#| fig-cap: 
#|   - "(a) Per-Item Variability"
#|   - "(b) Per-Participant Variability"
# (a)
ggplot(items) +
  geom_abline(aes(intercept = b_0+I0_i, slope=b_1, group=i_id)) +
  geom_abline(intercept=b_0, slope=b_1, colour="red") +
  xlim(1.5, 7.5) +
  ylim(4.5, 7.5)
# (b)
ggplot(participants) +
  geom_abline(aes(intercept = b_0+P0_p, slope=b_1, group=p_id)) +
  geom_abline(intercept=b_0, slope=b_1, colour="red") +
  xlim(1.5, 7.5) +
  ylim(4.5, 7.5)
```

Here, the red line shows the fixed effect, while the black lines show variability between (a) items and (b) participants. You can clearly see that there is less variability between items (`s_I0 =` `{r} s_I0`) than between participants (`s_P0 =` `{r} s_P0`).

## Simulating Trials

In a trial, a single participant responds to a single item. We simulate trials by simulating the meeting between participants and items.

As @debruineUnderstandingMixedEffectsModels2021 do, we here simulate this using the `crossing()` function from `tidyr`. This simulates a study in which all `{r} n_participants` participants respond to `{r} n_items` items, producing a total `{r} n_items * n_participants` trials. For each unique combination of item and participant, we also simulate the residuals, $e_{ip}$.

```{r}
# generate trials
trials <- crossing(items, participants) |>
  mutate(e_ip = rnorm(n(), 0, s_e))

# print the first few rows
print(trials)
```

::: callout-note
As well as simulating all possible combinations of items and participants, you could also simulate other combinations of participants and items. For example, you could simulate participants responding to N random trials, or you could create 10 separate stimulus lists, with 10% of your participants being assigned to each set of items. Similarly, you could use real data to inform your simulation, such as taking real items and real Zipf values from a frequency corpus.
:::

## Simulating Responses

Now that we have simulated trials, we can use our formula to simulate each participant's response to each item.

As a quick reminder, here is the formula again:

$$logRT_{ip} = \beta_0 + \beta_1 Zipf_{ip} + I0_i + P0_p + e_{ip}$$

We use this formula to add the simulated log response times to the `trials` dataframe:

```{r}
trials <- mutate(
    trials,
    log_rt = b_0 + b_1 * zipf + I0_i + P0_p + e_ip
  )
```

If we plot these observations coloured by items or participants, you can see how item and participant variability is reflected in the structure of the data:

```{r}
#| layout-ncol: 2
#| fig-height: 4
#| fig-width: 3
#| fig-cap: 
#|   - "(a) Trials coloured by Item"
#|   - "(b) Trials coloured by Participant"
# (a)
ggplot(trials, aes(zipf, log_rt, colour=factor(i_id))) +
  geom_point() +
  theme(legend.position="none")
# (b)
ggplot(trials, aes(zipf, log_rt, colour=factor(p_id))) +
  geom_point() +
  theme(legend.position="none")
```
