# Random Correlations

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false
source("ggplot2_setup.R", local=TRUE)
set.seed(1050109868)  # from sample(.Machine$integer.max, 1)
```

## What are Random Correlations?

In the [last section](03-random_slopes), we saw how we can simulate not only random variability between participants' *overall* response times, but also variability in the magnitude of slopes. We did this by modelling both $P_{0p}$ (random intercepts) and $P_{1p}$ (random slopes):

$$logRT_{ip} = \beta_0 + I_{0i} + P_{0p} + (\beta_1 + P_{1p}) Zipf_{ip} + e_{ip}$$

We simulated the random effects by drawing from independent normal distributions. However, this assumes that random intercepts and slopes are uncorrelated. It's not unreasonable to expect that there might actually be a correlation here. For example, people who respond more quickly may also show a larger effect of word frequency.

We can simulate these relationships by drawing $P_{0p}$ and $P_{1p}$ from multivariate normal distributions:

$$
\begin{pmatrix}
    P_{0p} \\ P_{1p}
    \end{pmatrix}
    \sim \mathcal{N}\!\left(\mathbf{0},\, \Sigma_P\right) \\
$$

Rather than a single standard deviation, as we've seen so far, $\Sigma_P$ is a covariance matrix, describing the standard deviations (variances) of $P_0$ and $P_1$ on the diagonal, and the covariances on the upper and lower triangle.

$$
\Sigma_P = 
    \begin{pmatrix}
    \sigma^2_{P0} & \sigma_{P01} \\
    \sigma_{P01} & \sigma^2_{P1}
    \end{pmatrix}
$$

Another way of writing the covariances is as correlations ($\rho$) multiplied by the standard deviations $\sigma_{P0}$ and $\sigma_{P1}$:

$$
\Sigma_P = 
    \begin{pmatrix}
    \sigma^2_{P0} & \rho_P \sigma_{P0} \sigma_{P1} \\
    \rho_P \sigma_{P0} \sigma_{P1} & \sigma^2_{P1}
    \end{pmatrix}
$$

We will continue to simulate the item random intercepts and the residuals as just independent normal distributions:

$$\begin{aligned}
I_{0i} &\sim \mathcal{N}(0, \sigma_{I0}) \\
e_{ip} &\sim \mathcal{N}(0, \sigma_e)
\end{aligned}$$

## Packages

We'll use the following packages in this section:

```{r}
#| label: packages
#| message: false
#| warning: false
library(lme4)  # for fitting LMMs
library(dplyr)  # for data wrangling
library(tidyr)  # for data wrangling
library(readr)  # for importing data
library(faux)  # for simulating multivariate normals
library(ggplot2)  # for plotting
```

## Simulating Multivariate Normals

To simulate multivariate normal distributions, we can no longer use `rnorm()`. Instead, we can use the `rnorm_multi()` function from `faux` [@debruineFauxSimulationFactorial2020].

For example, to simulate observations from two normally distributed variables, `a` and `b`, with a correlation of 0.76, we can do the following:

```{r}
# number of observations
n_ab <- 250

# means of the normal distributions
mu_a <- 7.2
mu_b <- 0

# SDs of the normal distributions
sigma_a <- 0.5
sigma_b <- 10

# correlation of the normal distributions
r_ab <- 0.76

# simulate observations
ab_df <- rnorm_multi(
  # number of observations
  n = n_ab,
  # number of variables
  vars = 2,
  varnames = c("a", "b"),
  # means
  mu = c(mu_a, mu_b),
  # SDs
  sd = c(sigma_a, sigma_b),
  # correlation
  r = r_ab
) |>
  as_tibble()

print(ab_df)
```

If we plot the simulated observations, we can indeed see that `a` and `b` are Gaussian and correlated:

```{r}
#| layout-ncol: 3
#| fig-height: 3
#| fig-width: 2
#| fig-cap: 
#|   - "(a) Density of a"
#|   - "(b) Density of b"
#|   - "(c) Relationship"
# (a)
ggplot(ab_df, aes(a)) +
  geom_density()
# (b)
ggplot(ab_df, aes(b)) +
  geom_density()
# (c)
ggplot(ab_df, aes(a, b)) +
  geom_point()
```
::: callout-note
Above, we passed a single correlation to the `r` argument. In cases where you have more than 2 variables, you can pass a full correlation matrix.

```{r}
#| eval: false
#| code-fold: true
#| code-summary: Show example
rnorm_multi(
  # number of observations
  n = 1000,
  # number of variables
  vars = 3,
  # means
  mu = c(0, 10, 4.2),
  # SDs
  sd = c(1, 7, 2.5),
  # correlation
  r = matrix(data=c(
    1, 0.5, 0.21,
    0.5, 1, 0.7,
    0.21, 0.7, 1
  ), nrow=3)
)
```
:::

## Defining Parameters

These are the parameters that I suggest for this simulation. These are the same as before, except now we also have the correlation between `s_P0` and `s_P1`, which I've called `r_P`. I've set this to -.74, meaning that participants with larger (slower) average response times should also have a reduced effect of word frequency.

```{r}
#| label: params
# Fixed-effect intercept and slope
# (same as the parameters in the section on linear models)
b_0 <- 6.85
b_1 <- -0.1

# SD of per-item random intercepts
s_I0 <- 0.08

# SD of per-participant random intercepts
s_P0 <- 0.16
s_P1 <- 0.04
r_P  <- -0.74

# residual SD
s_e <- 0.33
```

## Simulating Correlated Random Effects

Using the [`rnorm_multi()` example above](#simulating-multivariate-normals), write code to simulate the random intercepts and slopes for both items and participants. To ensure that the `lmer()` model will have enough information to estimate the random correlations, I also recommend increasing `n_participants`.

```{r}
#| code-fold: true
#| code-summary: "Show solution"
# simulate 50 items
n_items <- 50

items <- tibble(
  # item IDs
  i_id = 1:n_items,
  # item word frequencies
  zipf = runif(n_items, 1.5, 7.5),
  # item intercepts
  I_0i = rnorm(n_items, 0, s_I0)
)

# simulate 30 participants
n_participants <- 30

participants <- rnorm_multi(
  n = n_participants,
  vars = 2,
  varnames = c("P_0p", "P_1p"),
  mu = c(0, 0),
  sd = c(s_P0, s_P1),
  r = r_P
) |>
  mutate(p_id = 1:n_participants)
```

## Visualising Random Effects

If we now visualise the variability between items and participants, you can see that participants with faster response times also tend to have steeper slopes in the effect of Zipf:

```{r}
#| layout-ncol: 3
#| fig-height: 3
#| fig-width: 2
#| fig-cap: 
#|   - "(a) Per-Item Variability"
#|   - "(b) Per-Participant Variability"
#|   - "(c) Intercept-Slope Correlation"
# (a)
ggplot(items) +
  geom_abline(aes(intercept = b_0+I_0i, slope=b_1, group=i_id)) +
  geom_abline(intercept=b_0, slope=b_1, colour="red") +
  lims(x=c(1.5, 7.5), y=c(5, 7.5)) +
  labs(x="Zipf", y="Log RT")
# (b)
ggplot(participants) +
  geom_abline(aes(intercept = b_0+P_0p, slope=b_1+P_1p, group=p_id)) +
  geom_abline(intercept=b_0, slope=b_1, colour="red") +
  lims(x=c(1.5, 7.5), y=c(5, 7.5)) +
  labs(x="Zipf", y="Log RT")
# (c)
ggplot(participants, aes(b_0 + P_0p, b_1 + P_1p)) +
  geom_point() +
  geom_smooth(method="lm")
```

As before, the <span style="color:red;">**red line**</span> shows the fixed effect, while the **black lines** show variability between (a) items and (b) participants. The <span style="color:blue;">**blue line**</span> in the last panel shows the relationship between per-participant intercepts ($\beta_0 + P_{0p}$) and slopes ($\beta_0 + P_{1p}$).

## Simulating Trials and Responses

These sections are actually identical to what we did before. We get all combinations of items and participants, and then simulate:

* $e_{ip}$ (`e_ip`) as random noise of mean 0 and SD $\sigma_e$ (`s_e`)
* $logRT$ (`log_rt`) using the equation at the top

```{r}
# generate trials
trials <- crossing(items, participants) |>
  mutate(
    e_ip = rnorm(n(), 0, s_e),
    log_rt = b_0 + I_0i + P_0p + (b_1 + P_1p) * zipf + e_ip
  )

# print the first few rows
print(trials)
```

## Parameter Retrieval

To fit a model that can estimate the [simulated random correlation](#defining-parameters), we can now use one single bar (`|`) rather than double bars (`||`). This tells `lme4` to estimate and account for the influence of random correlations:

```{r}
#| code-fold: true
#| code-summary: "Show code"
#| output-fold: true
#| output-summary: "Show output"
m <- lmer(
  log_rt ~ 1 + zipf +
    (1 | i_id) +
    (1 + zipf | p_id),
  data = trials
)

summary(m)
```
