# Power Analyses

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false
source("ggplot2_setup.R", local=TRUE)
set.seed(1237865070)  # from sample(.Machine$integer.max, 1)
```

## Summary so far

Over the last few sections, we've built up code for simulating data from the following model:

$$\begin{aligned}
logRT_{ip} &= \beta_0 + I_{0i} + P_{0p} + (\beta_1 + P_{1p}) Zipf_{ip} + e_{ip} \\
I_{0i} &\sim \mathcal{N}(0, \sigma_{I0}) \\
\begin{pmatrix}
    P_{0p} \\ P_{1p}
    \end{pmatrix}
    &\sim \mathcal{N}\!\left(\mathbf{0},\, \Sigma_P\right) \\
\Sigma_P &= 
    \begin{pmatrix}
    \sigma^2_{P0} & \rho_P \sigma_{P0} \sigma_{P1} \\
    \rho_P \sigma_{P0} \sigma_{P1} & \sigma^2_{P1}
    \end{pmatrix} \\
e_{ip} &\sim \mathcal{N}(0, \sigma_e)
\end{aligned}$$

One of the ways in which we could apply this code is in power analyses. With a power analysis, we could ask:

> "Assuming our simulated coefficients are correct, what is the probability of correctly detecting the effect of word frequency?"

Using multi-level simulations is a powerful way of conducting power analyses. In this section, we'll build some code for conducting power analyses using the code we've built so far.

## Packages

We'll use the following packages in this section:

```{r}
#| label: packages
#| message: false
#| warning: false
library(lme4)  # for fitting LMMs
library(dplyr)  # for data wrangling
library(tidyr)  # for data wrangling
library(purrr)  # for iterating
library(readr)  # for importing data
library(faux)  # for simulating multivariate normals
library(lmerTest)  # for calculating significance
library(ggplot2)  # for plotting
```

## Modularising Code

To use the code we have come up with in a power analysis, we first want to *modularise* it. We want to write a self-contained function that takes our parameters as arguments, and returns simulated trials as an output

Functions in R have the following syntax:

```{r}
multiply <- function(a, b) {
  # calculate result
  res <- a * b
  # return the result by having the variable be the last line in the function
  res
}
```

Once a function is defined, we can call it to get the result:

```{r}
multiply(23, 2)
```

We can also define default arguments in the function. For example, if we think that most of the time people use our function, they will want to double a number, we could set the default for `b = 2`. Then, if people don't explicitly give a value for a, the function will use the default value of `2`.

```{r}
multiply <- function(a, b = 2) {
  res <- a * b
  res
}

multiply(4)
```

Using the code snippets from the [last section](04-random_correlations), write a function, `sim_trials()`, that takes all free parameters as inputs and returns the simulated trials. I recommend setting all arguments to defaults based on the numbers we've used so far.

```{r}
#| code-fold: true
#| code-summary: Show solution
sim_trials <- function(
    # all parameters with default values
    n_items = 50, n_participants = 30,
    b_0 = 6.85, b_1 = -0.1,
    s_I0 = 0.08,
    s_P0 = 0.16, s_P1 = 0.04,
    r_P = -0.74,
    s_e = 0.33
  ) {
  # simulate items
  items <- tibble(
    i_id = 1:n_items,
    zipf = runif(n_items, 1.5, 7.5),
    I_0i = rnorm(n_items, 0, s_I0)
  )
  
  # simulate participants
  participants <- rnorm_multi(
    n = n_participants,
    vars = 2,
    varnames = c("P_0p", "P_1p"),
    mu = c(0, 0),
    sd = c(s_P0, s_P1),
    r = r_P
  ) |>
    mutate(p_id = 1:n_participants)
  
  # generate trials
  trials <- crossing(items, participants) |>
    mutate(
      e_ip = rnorm(n(), 0, s_e),
      log_rt = b_0 + I_0i + P_0p + (b_1 + P_1p) * zipf + e_ip
    )
  
  # return the simulated trials
  trials
}
```

### Example Usage

If your function is similar to mine, then we can call it to simulate some example data. If we don't provide any inputs, it will use the defaults:

```{r}
sim_trials()
```

Or, we can provide specific inputs to change them from the defaults. For example, here we simulate 250 participants, and increase the SD of random intercepts:

```{r}
sim_trials(n_participants=250, s_P0=0.3)
```

## Hypothesis Testing

There are many ways to test hypotheses. Even within the null hypothesis significance testing approach, there are many ways to calculate a *p* value.

* **Wald *p* values** are the *p* values printed in the model `summary()` outputs for `lme4` models. Wald *p* values capture whether a model parameter is significantly different from zero.

* **Kenward-Roger corrected *p* values** are Wald *p* values, but with a correction to address inflated false-positive rates.

* **Likelihood ratio tests** provide *p* values for the significance of model comparisons. They capture whether data are significantly more likely to come from a model that includes the parameter of interest, relative to a simpler model that omits it.

[@lukeEvaluatingSignificanceLinear2017]
