# Power Analyses

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false
source("ggplot2_setup.R", local=TRUE)
set.seed(1237865072)  # from sample(.Machine$integer.max, 1)
```

## Summary so far

Over the last few sections, we've built up code for simulating data from the following model:

$$\begin{aligned}
logRT_{ip} &= \beta_0 + I_{0i} + P_{0p} + (\beta_1 + P_{1p}) Zipf_{ip} + e_{ip} \\
I_{0i} &\sim \mathcal{N}(0, \sigma_{I0}) \\
\begin{pmatrix}
    P_{0p} \\ P_{1p}
    \end{pmatrix}
    &\sim \mathcal{N}\!\left(\mathbf{0},\, \Sigma_P\right) \\
\Sigma_P &= 
    \begin{pmatrix}
    \sigma^2_{P0} & \rho_P \sigma_{P0} \sigma_{P1} \\
    \rho_P \sigma_{P0} \sigma_{P1} & \sigma^2_{P1}
    \end{pmatrix} \\
e_{ip} &\sim \mathcal{N}(0, \sigma_e)
\end{aligned}$$

One of the ways in which we could apply this code is in power analyses. With a power analysis, we could ask:

> "Assuming our simulated coefficients are correct, what is the probability of correctly detecting the effect of word frequency?"

Using multi-level simulations is a powerful way of conducting power analyses. In this section, we'll build some code for conducting power analyses using the code we've built so far.

## Packages

We'll use the following packages in this section:

```{r}
#| label: packages
#| message: false
#| warning: false
library(lme4)  # for fitting LMMs
library(dplyr)  # for data wrangling
library(tidyr)  # for data wrangling
library(purrr)  # for iterating
library(readr)  # for importing data
library(faux)  # for simulating multivariate normals
library(lmerTest)  # for calculating significance
library(broom.mixed)  # for getting tidy summaries from models
library(ggplot2)  # for plotting
```

## Modularising Code

To use the code we have come up with in a power analysis, we first want to *modularise* it. We want to write a self-contained function that takes our parameters as arguments, and returns simulated trials as an output

Functions in R have the following syntax:

```{r}
multiply <- function(a, b) {
  # calculate result
  res <- a * b
  # return the result by having the variable be the last line in the function
  res
}
```

Once a function is defined, we can call it to get the result:

```{r}
multiply(23, 2)
```

We can also define default arguments in the function. For example, if we think that most of the time people use our function, they will want to double a number, we could set the default for `b = 2`. Then, if people don't explicitly give a value for a, the function will use the default value of `2`.

```{r}
multiply <- function(a, b = 2) {
  res <- a * b
  res
}

multiply(4)
```

Using the code snippets from the [last section](04-random_correlations), write a function, `sim_trials()`, that takes all free parameters as inputs and returns the simulated trials. I recommend setting all arguments to defaults based on the numbers we've used so far.

As a last step, we should also [scale the Zipf variable](04-random_correlations#rescale-continuous-predictors) to promote model convergence.

```{r}
#| code-fold: true
#| code-summary: Show solution
sim_trials <- function(
    # all parameters with default values
    n_items = 50, n_participants = 30,
    b_0 = 6.85, b_1 = -0.1,
    s_I0 = 0.08,
    s_P0 = 0.16, s_P1 = 0.04,
    r_P = -0.74,
    s_e = 0.33
  ) {
  # simulate items
  items <- tibble(
    i_id = 1:n_items,
    zipf = runif(n_items, 1.5, 7.5),
    I_0i = rnorm(n_items, 0, s_I0)
  )
  
  # simulate participants
  participants <- rnorm_multi(
    n = n_participants,
    vars = 2,
    varnames = c("P_0p", "P_1p"),
    mu = c(0, 0),
    sd = c(s_P0, s_P1),
    r = r_P
  ) |>
    mutate(p_id = 1:n_participants)
  
  # generate trials
  trials <- crossing(items, participants) |>
    mutate(
      e_ip = rnorm(n(), 0, s_e),
      log_rt = b_0 + I_0i + P_0p + (b_1 + P_1p) * zipf + e_ip,
      # also include a scaled version of the zipf variable
      zipf_sc = scale(zipf)
    )
  
  # return the simulated trials
  trials
}
```

### Example Usage

If your function is similar to mine, then we can call it to simulate some example data. If we don't provide any inputs, it will use the defaults:

```{r}
sim_trials()
```

Or, we can provide specific inputs to change them from the defaults. For example, here we simulate 250 participants, and increase the SD of random intercepts:

```{r}
sim_trials(n_participants=250, s_P0=0.3)
```

## Hypothesis Testing

There are many ways to test hypotheses. Even within the null hypothesis significance testing approach, there are many ways to calculate a *p* value.

* **Wald *p* values** are the *p* values printed in the model `summary()` outputs for `lme4` models. They correspond to *z* values calculated as $\beta/SE$. The *p* values then reflect the probability of seeing a *z* value at least this large, assuming the null hypothesis is true.

* **Satterthwaite and Kenward-Roger *p* values** are Wald *p* values, but with a correction to the degrees of freedom to address inflated false-positive rates.

* **Likelihood ratio tests** provide *p* values for the significance of model comparisons. They capture whether data are significantly more likely to come from a model that includes the parameter of interest, relative to a simpler model that omits it. They reflect the probability of seeing a likelihood ratio at least as large as that observed, assuming that the models actually explain the data equally well.

These approaches differ in their power and false-positive rates. Satterthwaite and Kenward-Roger *p* values are generally the best calibrated tests, while Wald *p* values result in high false-positive rates [@lukeEvaluatingSignificanceLinear2017].

## Convergence Issues

Let's generate a dataset using the default parameters:

```{r}
trials <- sim_trials()
```

Next let's fit a model with our full ("maximal") mixed-effects formula:

```{r}
m <- lmer(
  log_rt ~ 1 + zipf_sc +
    (1 | i_id) +
    (1 + zipf_sc | p_id),
  data = trials
)
```

The model converges nicely. We can confirm this using an approach [we saw briefly in the last section](04-random_correlations#parameter-retrieval):

```{r}
model_converged <- is.null(m@optinfo$conv$lme4$messages)

print(model_converged)
```

When running power simulations, we can use this to programmatically address convergence issues. Addressing any convergence issues in this way is valid if the rules we use to address convergence issues in the simulations are the same rules we actually plan to use in the eventual study.

## Getting *p* Values

If you've loaded the `lmerTest` package, then the model summary will print tests for the fixed effects with degrees of freedom and p values.

```{r}
summary(m)
```

We can get these results in a tidier format using the `broom.mixed::tidy()` function:

```{r}
tidy(m, effects="fixed", ddf.method="Satterthwaite")
```

## Running Power Simulations

We now have everything we need to run some simulations to calculate power. Let's calculate the statistical power to detect the effect of Zipf when we have 50 items and 50 participants, but the frequency effect is around 3 times smaller than we originally simulated. To make sure it finishes in a reasonable time, we'll just run 250 simulations.

```{r}
#| warning: false
#| message: false
n_sims <- 250

sims_res <- lapply(1:n_sims, function(i) {
  # simulate data
  trials <- sim_trials(n_items=50, n_participants=50, b_1=-0.033) |>
    mutate(zipf_sc = scale(zipf))
  
  # fit model
  m <- lmer(
    log_rt ~ 1 + zipf_sc +
      (1 | i_id) +
      (1 + zipf_sc | p_id),
    data = trials
  )
  
  # detect whether converged (TRUE if converged)
  m_converged <- is.null(m@optinfo$conv$lme4$messages)
  
  # if not converged, refit with random correlations removed
  if (!m_converged) {
    removed_rcs <- TRUE  # record that the random correlations were removed
    
    m <- lmer(
      log_rt ~ 1 + zipf_sc +
        (1 | i_id) +
        (1 + zipf_sc || p_id),
      data = trials
    )
  } else {
    removed_rcs <- FALSE  # record that the random correlations were kept
  }
  
  # get tidy summary with Satterthwaite p values
  tidy(m, effects="fixed", ddf.method="Satterthwaite") |>
    # only get the results for the zipf parameter (not also the intercept)
    filter(term == "zipf_sc") |>
    mutate(
      # store the simulation number
      i = i,
      # store whether converged
      converged = is.null(m@optinfo$conv$lme4$messages),
      # store whether random correlations were removed
      removed_rcs = removed_rcs
    )
})
```

This will give us a list of dataframes (one dataframe for each simulation). We can use `purrr::list_rbind()` to join these into one large dataframe.

```{r}
sims_res_df <- list_rbind(sims_res)

sims_res_df
```

We can also see that random correlations were removed from about 1 third of models because of convergence issues.

```{r}
count(sims_res_df, removed_rcs)
```

As you can see here, all of the models then converged. This means that removing the random correlations addressed the nonconvergence in those original 5 models.

```{r}
count(sims_res_df, converged)
```

## Calculating Power

We can estimate the power as the proportion of simulations in which our p value is smaller than alpha.

```{r}
mean(sims_res_df$p.value < 0.05)
```

This tells us that if we present 50 participants with 50 items, and the effect of Zipf is -0.033, we should expect 94.8% of studies to correctly detect the effect using our analysis method.

## Parallelisation

Running large numbers of simulations, in which each simulation requires fitting a mixed-effects model, is quite intensive in terms of computation and time. Since R is fitting these models in a single-thread manner, we could make this much faster by parallelising our simulations across cores.
