# Power Analyses

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false
source("ggplot2_setup.R", local=TRUE)
set.seed(1237865072)  # from sample(.Machine$integer.max, 1)
```

## Summary so far

Over the last few sections, we've built up code for simulating data from the following model:

$$\begin{aligned}
logRT_{ip} &= \beta_0 + I_{0i} + P_{0p} + (\beta_1 + P_{1p}) Zipf_{ip} + e_{ip} \\\\
I_{0i} &\sim \mathcal{N}(0, \sigma_{I0}) \\\\
\begin{pmatrix}
    P_{0p} \\ P_{1p}
    \end{pmatrix}
    &\sim \mathcal{N}\!\left(\mathbf{0},\, \Sigma_P\right) \\\\
\Sigma_P &= 
    \begin{pmatrix}
    \sigma^2_{P0} & \rho_P \sigma_{P0} \sigma_{P1} \\
    \rho_P \sigma_{P0} \sigma_{P1} & \sigma^2_{P1}
    \end{pmatrix} \\\\
e_{ip} &\sim \mathcal{N}(0, \sigma_e)
\end{aligned}$$

One of the ways in which we could apply this code is in power analyses. With a power analysis, we could ask:

> "Assuming our simulated coefficients are correct, what is the probability of correctly detecting the effect of word frequency?"

Using multi-level simulations is a powerful way of conducting power analyses. In this section, we'll build some code for conducting power analyses using the code we've built so far.

## Packages

We'll use the following packages in this section:

```{r}
#| label: packages
#| message: false
#| warning: false
library(lme4)  # for fitting LMMs
library(dplyr)  # for data wrangling
library(tidyr)  # for data wrangling
library(purrr)  # for iterating
library(faux)  # for simulating multivariate normals
library(lmerTest)  # for calculating significance
library(broom.mixed)  # for getting tidy summaries from models
library(parallel)  # for parallelisation
library(ggplot2)  # for plotting
```

## Modularising Code

To use the code we have come up with in a power analysis, we first want to *modularise* it. We want to write a self-contained function that takes our parameters as arguments, and returns simulated trials as an output

Functions in R have the following syntax:

```{r}
multiply <- function(a, b) {
  # calculate result
  res <- a * b
  # return the result by having the variable be the last line in the function
  res
}
```

Once a function is defined, we can call it to get the result:

```{r}
multiply(23, 2)
```

We can also define default arguments in the function. For example, if we think that most of the time people use our function, they will want to double a number, we could set the default for `b = 2`. Then, if people don't explicitly give a value for a, the function will use the default value of `2`.

```{r}
multiply <- function(a, b = 2) {
  res <- a * b
  res
}

multiply(4)
```

Using the code snippets from the [last section](04-random_correlations), write a function, `sim_trials()`, that takes all free parameters as inputs and returns the simulated trials. I recommend setting all arguments to defaults based on the numbers we've used so far.

As a last step, the function should also [scale the Zipf variable](04-random_correlations#rescale-continuous-predictors) to promote model convergence.

```{r}
#| code-fold: true
#| code-summary: Show solution
sim_trials <- function(
    # all parameters with default values
    n_items = 50, n_participants = 30,
    b_0 = 6.85, b_1 = -0.1,
    s_I0 = 0.08,
    s_P0 = 0.16, s_P1 = 0.04,
    r_P = -0.74,
    s_e = 0.33
  ) {
  # simulate items
  items <- tibble(
    i_id = 1:n_items,
    zipf = runif(n_items, 1.5, 7.5),
    I_0i = rnorm(n_items, 0, s_I0)
  )
  
  # simulate participants
  participants <- rnorm_multi(
    n = n_participants,
    vars = 2,
    varnames = c("P_0p", "P_1p"),
    mu = c(0, 0),
    sd = c(s_P0, s_P1),
    r = r_P
  ) |>
    mutate(p_id = 1:n_participants)
  
  # generate trials
  trials <- crossing(items, participants) |>
    mutate(
      e_ip = rnorm(n(), 0, s_e),
      log_rt = b_0 + I_0i + P_0p + (b_1 + P_1p) * zipf + e_ip,
      # also include a scaled version of the zipf variable
      zipf_sc = scale(zipf)
    )
  
  # return the simulated trials
  trials
}
```

### Example Usage

If your function is similar to mine, then we can call it to simulate some example data. If we don't provide any inputs, it will use the defaults:

```{r}
sim_trials()
```

Or, we can provide specific inputs to change them from the defaults. For example, here we simulate 250 participants, and increase the SD of random intercepts:

```{r}
sim_trials(n_participants=250, s_P0=0.3)
```

## Hypothesis Testing

There are many ways to test hypotheses. Even within the null hypothesis significance testing approach, there are many ways to calculate a *p* value.

- **Wald *p* values** are the *p* values printed in the model `summary()` outputs for `lme4` models. They correspond to *z* values calculated as $\beta/SE$. The *p* values then reflect the probability of seeing a *z* value at least as extreme as that observed, assuming the null hypothesis is true.

- **Satterthwaite and Kenward-Roger *p* values** are Wald *p* values, but with a correction to the degrees of freedom to address inflated false-positive rates.

- **Likelihood ratio tests** provide *p* values for the significance of model comparisons. They capture whether data are significantly more likely to come from a model that includes the parameter of interest, relative to a simpler model that omits it. They reflect the probability of seeing a likelihood ratio at least as large as that observed, assuming that the models actually explain the data equally well.

These approaches differ in their power and false-positive rates. Satterthwaite and Kenward-Roger *p* values are generally the best calibrated tests, while Wald *p* values result in high false-positive rates [@lukeEvaluatingSignificanceLinear2017].

## Convergence Issues

Let's generate a dataset using the default parameters:

```{r}
trials <- sim_trials()
```

Next let's fit a model with our full ("maximal") mixed-effects formula:

```{r}
m <- lmer(
  log_rt ~ 1 + zipf_sc +
    (1 | i_id) +
    (1 + zipf_sc | p_id),
  data = trials
)
```

The model converges nicely. We can confirm this using an approach [we saw briefly in the last section](04-random_correlations#parameter-retrieval):

```{r}
model_converged <- is.null(m@optinfo$conv$lme4$messages)

print(model_converged)
```

When running power simulations, we can use this to programmatically address convergence issues. Addressing any convergence issues in this way is valid if the rules we use to address convergence issues in the simulations are the same rules we actually plan to use in the eventual study.

## Getting *p* Values

If you've loaded the `lmerTest` package, then the model summary will print tests for the fixed effects with degrees of freedom and p values.

```{r}
summary(m)
```

We can get these results in a tidier format using the `broom.mixed::tidy()` function:

```{r}
tidy(m, effects="fixed", ddf.method="Satterthwaite")
```

## Running Power Simulations

We now have everything we need to run some simulations to calculate power. Let's calculate the statistical power to detect the effect of Zipf when we have 50 items and 50 participants, but the frequency effect is around 3 times smaller than we originally simulated. To make sure it finishes in a reasonable time, we'll just run 250 simulations.

```{r}
#| warning: false
#| message: false
n_sims <- 250

# function for running a single simulation and returning the results
sim_res <- function(sim_nr, options = list()) {
  # simulate data (also pass the list of options)
  trials <- do.call(sim_trials, options)
  
  # fit model
  m <- lmer(
    log_rt ~ 1 + zipf_sc +
      (1 | i_id) +
      (1 + zipf_sc | p_id),
    data = trials
  )
  
  # detect whether converged (TRUE if converged)
  m_converged <- is.null(m@optinfo$conv$lme4$messages)
  
  # if not converged, refit with random correlations removed
  if (!m_converged) {
    removed_rcs <- TRUE  # record that the random correlations were removed
    
    m <- lmer(
      log_rt ~ 1 + zipf_sc +
        (1 | i_id) +
        (1 + zipf_sc || p_id),
      data = trials
    )
  } else {
    removed_rcs <- FALSE  # record that the random correlations were kept
  }
  
  # get tidy summary with Satterthwaite p values
  res <- tidy(m, effects="fixed", ddf.method="Satterthwaite") |>
    # only get the results for the zipf parameter (not also the intercept)
    filter(term == "zipf_sc") |>
    mutate(
      # store the simulation number
      sim_nr = sim_nr,
      # store whether converged
      converged = is.null(m@optinfo$conv$lme4$messages),
      # store whether random correlations were removed
      removed_rcs = removed_rcs
    )
  
  # add the non-default arguments as columns in the output
  options_df <- as_tibble(options)
  res <- bind_cols(res, options_df)
  
  # return the dataframe
  res
}

# run the simulations, passing the options as arguments to the function
sims <- lapply(
  # pass the simulation numbers as the first argument
  1:n_sims, sim_res,
  # then pass the desired options as subsequent arguments
  options = list("n_items"=50, "n_participants"=50, "b_1"=-0.033)
)
```

This will give us a list of dataframes (one dataframe for each simulation). We can use `purrr::list_rbind()` to join these into one large dataframe.

```{r}
sims_df <- list_rbind(sims)

sims_df
```

We can also see that random correlations were removed from a few models to address convergence issues.

```{r}
count(sims_df, removed_rcs)
```

We can also see that removing correlations from these 5 models did address convergence issues, because in the final check, all models converged.

```{r}
count(sims_df, converged)
```

## Calculating Power

We can estimate the power as the proportion of simulations in which our p value is smaller than alpha.

```{r}
mean(sims_df$p.value < 0.05)
```

This tells us that if we present 50 participants with 50 items, and the effect of Zipf is `{r} sims_df$b_1[[1]]`, then in the long run we should expect `{r} round(mean(sims_df$p.value < 0.05) * 100, 2)`% of studies to correctly detect the effect using our analysis method.

Given that we expect frequency to have a negative effect (i.e., speed up responses), we could also calculate the one-tailed power for the same alpha level:

```{r}
mean(sims_df$p.value < 0.1 & sims_df$estimate < 0)
```

Although they have the same false-positive rate, one-tailed significance tests are higher.

## Parallelisation

Running large numbers of simulations, in which each simulation requires fitting a mixed-effects model, is quite intensive in terms of computation and time. This is especially true if we want to run simulations across multiple combinations of parameters.

For instance, we often want to run simulations across multiple values of *N*, to identify the number of participants and/or items required to reach a certain power target. Let's run 250 simulations for each value of `n_participants` from 3 to 15.

```{r}
n_sims <- 250  # per combination of parameters
n_participants <- 3:15
```

Since R is fitting these models in a single-thread manner, we could make this much faster by parallelising our simulations across cores.

First, we build a dataframe that contains all combinations of parameters for the simulations we want to run. To do this, I use `expand_grid()`.

```{r}
sim_pars_df <- expand_grid(
  sim_nr = 1:n_sims,
  n_participants = n_participants
)
```

It's also important to ensure reproducibility. One approach is to pre-assign random seeds for each simulation:

```{r}
# assign random seeds for each simulation
max_int <- 2147483647  # .Machine$integer.max
total_n_sims <- nrow(sim_pars_df)  # get the total number of simulations to run
sim_pars_df <- mutate(sim_pars_df, seed = sample.int(max_int, total_n_sims))
```

We then want to split the dataframe we've built into a list that we can iterate over easily. Here we split them into single-row dataframes with the parameters for individual simulations.

```{r}
sim_pars <- group_split(sim_pars_df, sim_id = row_number())
```

Each item in `sim_pars` now contains the parameters for running one simulation, e.g.:

```{r}
sim_pars[[1]]
```

We also need to now initialise our cluster. Here I create a cluster of as many cores as the `parallel` library can detect on my machine:

```{r}
# get the number of cores
n_cores <- detectCores()
# create a cluster with that number of cores
cl <- makeCluster(n_cores)
```

We also need to make sure that each worker in the cluster has all the required objects and libraries loaded. Here we tell each worker to load the libraries that we use in the `sim_trials()` and `sim_res()` functions.

```{r}
cl_pkg <- clusterEvalQ(cl, {
  library(dplyr)
  library(tidyr)
  library(lme4)
  library(lmerTest)
  library(faux)
  library(broom.mixed)
})
```

The clusters also need to have the `sim_trials()` and `sim_res()` functions available to them, so we pass the function objects.

```{r}
cl_obj <- clusterExport(cl, list("sim_trials", "sim_res"))
```

Finally, we can run the simulations.

```{r}
sims <- parLapply(cl, sim_pars, function(pars) {
  # set the seed for reproducibility
  set.seed(pars$seed)
  # run the simulations, passing n_participants via the options list
  sim_res(
    sim_nr = pars$sim_nr,
    options = list("n_participants" = pars$n_participants)
  )
})
```

## Plotting Power over Parameters

Again, we can use `list_rbind()` to convert the list of outputs to a single dataframe. We also add a column, `is_sig`, which will contain information on whether each individual simulation had a significant effect of Zipf.

```{r}
sims_df <- list_rbind(sims) |>
  mutate(is_sig = as.integer(p.value < 0.05))
```

Here we calculate the power for each number of participants. I also calculate a binomial 95% confidence interval for each value of `n_participants`.

```{r}
sims_summ <- sims_df |>
  group_by(n_participants) |>
  summarise(
    power = mean(is_sig),
    ci_lwr = binom.test(sum(is_sig), n(), conf.level=0.95)$conf.int[1],
    ci_upr = binom.test(sum(is_sig), n(), conf.level=0.95)$conf.int[2]
  )
```

We can now plot how expected statistical power varies as we change the number of participants. Let's also add a horizontal line to show where 90% power would be:

```{r}
#| fig-height: 3
#| fig-width: 4
ggplot(sims_summ, aes(n_participants, power)) +
  geom_point() +
  geom_linerange(aes(ymin=ci_lwr, ymax=ci_upr)) +
  geom_line() +
  geom_hline(yintercept=0.9, linetype="dashed")
```

## Modelling Statistical Power

We can also go one step further. Rather than estimating power for each value of N Participants separately, we can fit a general linear model that includes N Participants as a coefficient. A model formula that tends to predict power quite well is `is_sig ~ 1 + sqrt(n_participants)`.

We could use such a model to make predictions about the expected power for a given value of `n_participants`. This can be very useful since, rather than running hundreds of simulations at each value of N, we can run fewer simulations in total, and then interpolate across the observed combinations to improve the precision of our predictions.

```{r}
# fit a model predicting power
power_mod <- glm(
  is_sig ~ 1 + sqrt(n_participants),
  family = binomial("logit"),
  data = sims_df
)

# predict power for a given number of participants
predict(
  power_mod,
  newdata = tibble(n_participants = 5),
  type = "response"
)
```

We could even predict the number of participants required to reach a certain power target. For example, if we want to reach 90% power, we can rearrange the formula for a logit-link binomial model to calculate the value of N associated with a probability of 0.9.

Since our [logit-link binomial model](https://en.wikipedia.org/wiki/Logistic_regression) for power is estimating

$$ log\left( \frac{p}{1-p} \right) = \beta_0 + \beta_1\sqrt{NParticipants}, $$

we can rearrange this to tell us the value of $NParticipants$ for a given $p$:

::: {.callout-note collapse="true"}
#### Show Rearrangement Details

$$\begin{aligned}
log\left( \frac{p}{1-p} \right) &= \beta_0 + \beta_1\sqrt{NParticipants} \\\\
log\left( \frac{p}{1-p} \right) - \beta_1\sqrt{NParticipants} &= \beta_0 \\\\
\beta_1\sqrt{NParticipants} &= log\left( \frac{p}{1-p} \right) - \beta_0 \\\\
\sqrt{NParticipants} &= \frac{ log\left( \frac{p}{1-p} \right) - \beta_0 }{\beta_1} \\\\
NParticipants &= \left( \frac { log\left( \frac{p}{1-p} \right) - \beta_0 }{ \beta_1 } \right) ^2
\end{aligned}$$
:::

$$NParticipants = \left( \frac { log\left( \frac{p}{1-p} \right) - \beta_0 }{ \beta_1 } \right) ^ 2$$

```{r}
power_target <- 0.9

logit_target <- log(power_target/(1-power_target))

power_b_0 <- coef(power_mod)[["(Intercept)"]]
power_b_1 <- coef(power_mod)[["sqrt(n_participants)"]]

target_n_sqrt <- (logit_target - power_b_0) / power_b_1

target_n <- target_n_sqrt ** 2

target_n
```

Based on this, we might plan to collect from 5 participants to reach at least `{r} round(power_target * 100, 2)`% power.

Finally, we can plot this model's predictions using `geom_smooth()`. I've added dashed lines at the `{r} round(power_target * 100, 2)`% power target, and the number of participants required to reach this target.

```{r}
#| fig-height: 3
#| fig-width: 4
ggplot(sims_summ, aes(n_participants, power)) +
  geom_smooth(
    aes(x=n_participants, y=is_sig),
    data = sims_df,
    method = glm,
    method.args = list("family" = binomial("logit")),
    formula = y ~ 1 + sqrt(x)
  ) +
  geom_point() +
  geom_hline(yintercept=power_target, linetype="dashed") +
  geom_vline(xintercept=target_n, linetype="dashed") +
  labs(x = "N Participants", y = "Statistical Power")
```

## Exercises

1. The power above was calculated using two-tailed power. Compare the power curves for one-tailed and two-tailed power. (Note: this should not require any new simulations!)

```{r}
#| code-fold: true
#| message: false
#| warning: false
#| code-summary: "Show solution"
#| layout-ncol: 2
#| fig-height: 3
#| fig-width: 4
# calculate significance for one- and two-tailed separately
sims_df <- list_rbind(sims) |>
  expand_grid(sig_test = c("one-tailed", "two-tailed")) |>
  mutate(is_sig = ifelse(
    sig_test == "one-tailed",
    as.integer(p.value < 0.1 & estimate < 0),
    as.integer(p.value < 0.05)
  ))

# calculate mean for one- and two-tailed separately
sims_summ <- sims_df |>
  group_by(n_participants, sig_test) |>
  summarise(power = mean(is_sig))

# plot
ggplot(sims_summ, aes(n_participants, power, colour=sig_test, group=sig_test)) +
  geom_smooth(
    aes(x = n_participants, y = is_sig),
    data = sims_df,
    method = glm,
    method.args = list("family" = binomial("logit")),
    formula = y ~ 1 + sqrt(x)
  ) +
  geom_point() +
  geom_hline(yintercept=power_target, linetype="dashed") +
  labs(x = "N Participants", y = "Statistical Power")
```

2. Run new simulations to consider possible values of both N Participants and N Items. I suggest 100 simulations per each unique combination of parameters. If possible, try to recreate the plot below:

```{r}
#| code-fold: true
#| message: false
#| warning: false
#| code-summary: "Show solution"
#| fig-height: 3
#| fig-width: 4
# set up parameters (16900 simulations in total!)
sim_pars_df <- expand_grid(
  sim_nr = 1:100,
  n_participants = 3:15,
  n_items = 3:15
)

# convert to list with parameters for each simulation
sim_pars <- group_split(sim_pars_df, sim_id = row_number())

# run simulations
sims <- parLapply(cl, sim_pars, function(pars) {
  set.seed(pars$seed)
  
  sim_res(
    sim_nr = pars$sim_nr,
    options = list(
      "n_participants" = pars$n_participants,
      "n_items" = pars$n_items
    )
  )
})

# convert to dataframe
sims_df <- list_rbind(sims) |>
  mutate(is_sig = as.integer(p.value < 0.05))

# calculate power per combination
sims_summ <- sims_df |>
  group_by(n_participants, n_items) |>
  summarise(power = mean(is_sig))

# plot
ggplot(sims_summ, aes(n_participants, power, colour=n_items, group=n_items)) +
  geom_point() +
  geom_smooth(
    aes(x = n_participants, y = is_sig),
    data = sims_df,
    method = glm,
    method.args = list("family" = binomial("logit")),
    formula = y ~ 1 + sqrt(x)
  ) +
  scale_colour_viridis_c()
```

3. The lines in the plot above actually fit used a separate model for each value of `n_items`. Can you think of a way of fitting one model that uses both `n_items` and `n_participants` as predictors? If possible, see whether you can produce the two plots below.

```{r}
#| code-fold: true
#| message: false
#| warning: false
#| code-summary: "Show solution"
#| fig-height: 3
#| fig-width: 3.5
#| layout-ncol: 2
# fit model with interaction between sqrt(N participant) and sqrt(N item)
power_mod <- glm(
  is_sig ~ 1 + sqrt(n_participants) * sqrt(n_items),
  family = binomial("logit"),
  data = sims_df
)

# get predicted power for each combination of Ns
combs_df <- expand_grid(n_participants=3:15, n_items=3:15)

combs_df$pred_power <- predict(
  power_mod,
  newdata = combs_df,
  type = "response"
)

# plot per-cell means
ggplot(sims_summ, aes(n_participants, n_items, fill=power)) +
  geom_raster() +
  scale_fill_viridis_c(limits = c(0, 1)) +
  labs(title = "Empirical Means")

# plot model predictions
ggplot(combs_df, aes(n_participants, n_items, fill=pred_power)) +
  geom_raster() +
  scale_fill_viridis_c(limits = c(0, 1)) +
  labs(title = "Model Predictions")
```
